#Behavioral delineation 

**Pronghorn Behavior**

To seperate out movement beahviors from GPS tracks, a hidden marrkov model can be used  
to parse out *encamped* **(1)** vs *exploratory* **(2)** behavior. 

First, we need to prep the data. I am using an HMM previously fitted using the *momentuHMM* package. I will use this data to fit net square displacement plots to see who is migrant, who is not, and if there are any other behaviors I can observe.

```{r include=FALSE}
# Load in packages ----
library(amt)
library(dplyr)
library(geosphere)
library(sf)
library(ggplot2)
library(lubridate)
library(gridExtra)

# Source function
source("/Users/vermanica/Documents/Projects/Analysis/scripts/Pronghorn/prep_fun.R")

#Set options----
options(stringsAsFactors = FALSE)
Sys.setenv(TZ = "UTC")

# 1. Load in data ----
ph <- readRDS("../../../Data/PH_data/processed_data/HMM_ph_states.rds")
class(ph)
str(ph)

# 2. Run function to prep data ----
nsd_prep(ph)

# 3. filter out year ---
# want range to be between 02/01/year - 02/01/ next year
data <- ph %>%
  select(XLOC = x,
         YLOC = y,
         dt = time,
         ID,
         state) %>%
  mutate(state = factor(state, levels = c(1, 2))) %>% 
  mutate(dt = ymd_hms(dt)) %>%
  filter(dt >= "2018-02-01" & dt <= "2019-02-03")  


str(data)
```

Next, we will fit NSD plots using an HMM output that gives us a breakdown of movements between "encamped" and "exploratory".

```{r include=FALSE}
# 4. Calculate NSD  ----
# run function to set this up
next_nsd(data)

# loop over indiv 
# want this grouped by indiv
individuals <- unique(data$ID)
# empty data frame for storage
data2 <- data.frame()

# nested for loop to loop NSD calculation over entire data set
for(i in 1:length(individuals)){
  # print status
  print(paste(i, individuals[i]))
  # temporary file where I can filter based on indiv to get indiv calculations
  temp <- data %>% 
    filter(ID == individuals[i])
  # for loop to loop calculations over temp data frame
  if(nrow(temp) > 2){
    for(j in 2:(nrow(temp) -1)){
      # first calculations for each indiv 
      temp$Xdist[1] <- temp$XLOC[2] - temp$XLOC[1]
      temp$Ydist[1] <- temp$YLOC[2] - temp$YLOC[1]
      temp$NSD[1]   <- as.integer(sqrt(temp$Xdist[1]^2+temp$Ydist[1]^2))
      # subsequent calculations for rest of the data per indiv
    temp$Xdist[j] <- temp$XLOC[j+1] - temp$XLOC[j] + temp$Xdist[j-1]
    temp$Ydist[j] <- temp$YLOC[j+1] - temp$YLOC[j] + temp$Ydist[j-1]
    temp$NSD[j] <- as.numeric(sqrt(temp$Xdist[j]^2 + temp$Ydist[j]^2))
    }
  }else{
    next
  }
  # combine the two data frames
    data2 <- rbind(data2, temp)
}
     
```
Finally, I will subset this data to three individuals presenting different movement patterns.

```{r include=TRUE}
# select idividuals presenting different movements
data1 <- data2 %>% filter(ID == "PR17F0015")
data3 <- data2 %>% filter(ID == "PR17F0012")
data4 <- data2 %>% filter(ID == "PR17F0033")

# plot each individual
p1 <- ggplot(data1, aes(x = dt, y = NSD, color = state)) +
                 geom_point() +
            expand_limits(y = c(0, 35000)) +
             theme_minimal()+
    xlab("Date") +
        ggtitle("PR17F0015 - Migrant")

p2 <- ggplot(data3, aes(x = dt, y = NSD, color = state)) +
                 geom_point() +
            expand_limits(y = c(0, 35000)) +
             theme_minimal()+
    xlab("Date") +
        ggtitle("PR17F0012 - Dispersal")

p3 <- ggplot(data4, aes(x = dt, y = NSD, color = state)) +
                 geom_point() +
            expand_limits(y = c(0, 35000)) +
             theme_minimal()+
    xlab("Date") +
        ggtitle("PR17F0033 - Unknown")

# group plots together
grid.arrange(p1,
  arrangeGrob(p2, p3, ncol = 2),
             nrow = 2)
```
These individuals are presenting different movements. The top plot shows a migrant individual, the bottom right shows a disperser, and the bottom left shows an individual whose movements are not able to be catagotized.

